Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-08-08 resubmission/02-antecedents.qmd
jc review

Recent research has highlighted the importance of disaggregating the concept of "meritocratic beliefs," particularly by distinguishing between meritocratic preferences and meritocratic perceptions [@castillo_multidimensional_2023]. While preferences refer to normative ideals about how rewards should be allocated (i.e., based on merit), perceptions refer to subjective evaluations of how meritocracy actually operates in practice (i.e., how rewards are allocated) [@janmaat_subjective_2013]. This distinction allows for a more precise understanding of how individuals make sense of inequality: people may endorse meritocracy normatively while simultaneously recognizing its failure in practice. In stratified societies, the gap between preferences and perceptions is particularly evident, as individuals may continue to uphold the ideal of meritocracy despite growing awareness of unequal opportunities [@mijs_belief_2022].

@@

Meritocratic perceptions function as a normative framework that legitimizes unequal outcomes, particularly when access to social goods is stratified by income or socioeconomic background. Prior studies have consistently shown that individuals who perceive society as meritocratic express lower support for redistribution [@tejero-peregrina_perceived_2025; @hoyt_mindsets_2023], greater legitimization of class inequality [@batruch_belief_2023], increased tolerance for inequality [@day_movin_2017], and lower perceptions of income inequality itself [@castillo_percepcion_2019]. In the regard of market justice, prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_socialization_2024]. Taken together, and as @busemeyer_positive_2021 argues, meritocratic narratives can operate as feedback mechanisms that shape public opinion and individual well-being by framing welfare outcomes as either deserved or undeserved. This psychological mechanism underscores the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal understandings of success and failure.

@@

Concerning public opinion, the ELSOC survey shows that Chile has a low level of agreement with the idea that it is fair for those with higher incomes to buy better welfare in the areas of education, healthcare, and pensions, with average support of 8.5% between 2016 and 2023. At the same time, around 55% and 47% of the public between 2016 and 2023 *disagreed* that effort and talent are rewarded accordingly. This coincides with 90% of citizens *agreeing* that income differences in Chile are *too large* [@castillo_changes_2025].

In sum, Chile represents an illustrative case of a particular form of *moral economy* [@koos_moral_2019], in which national socioeconomic characteristics, institutional arrangements, and public attitudes coexist. This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system [@madariaga_three_2020]. At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services [@somma_no_2021]. For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.

@@

Building upon the previous literature, this study proposes that attitudes toward market justice are shaped by a dynamic interplay between individuals' perceptions of economic inequality and their perceptions of meritocracy. Specifically, we argue that both perceptions independently and interactively influence the extent to which individuals endorse market-based distributions of social goods and services in Chile.

First, consistent with previous findings about redistributive preferences [@castillo_perception_2022], we expect that a higher perception of economic inequality will be associated with lower market justice preferences. When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay. Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution. This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force in shaping public attitudes toward inequality and justice [@canalesceron_sujeto_2021; @mac-clure_justicia_2024].

@@

Additionally, this study examines whether these relationships vary over time, particularly in light of broader social and political transformations. We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points. While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disipavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research. Instead, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.

>>>>

Recent research has highlighted the importance of disentangling the concept of "meritocratic beliefs," particularly by distinguishing between meritocratic preferences and meritocratic perceptions [@castillo_multidimensional_2023]. While preferences refer to normative ideals about how rewards should be allocated (i.e., based on merit), perceptions refer to subjective evaluations of how meritocracy actually operates in practice (i.e., how rewards are allocated) [@janmaat_subjective_2013]. This distinction allows for a more precise understanding of how individuals make sense of inequality: people may endorse meritocracy normatively while simultaneously recognizing its failure in practice. In stratified societies, the gap between preferences and perceptions is particularly evident, as individuals may continue to uphold the ideal of meritocracy despite growing awareness of unequal opportunities [@mijs_belief_2022].

@@

Meritocratic perceptions function as a normative framework that legitimizes unequal outcomes, particularly when access to social goods is stratified by income or socioeconomic background. Prior studies have consistently shown that individuals who perceive society as meritocratic express lower support for redistribution [@tejero-peregrina_perceived_2025; @hoyt_mindsets_2023], greater legitimization of class inequality [@batruch_belief_2023], increased tolerance for inequality [@day_movin_2017], and lower perceptions of income inequality itself [@castillo_percepcion_2019]. Regarding market justice preferences, prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_socialization_2024]. Taken together, and as @busemeyer_positive_2021 argues, meritocratic narratives can operate as feedback mechanisms that shape public opinion and individual well-being by framing welfare outcomes as either deserved or undeserved. Such mechanism underscores the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal understandings of success and failure.

@@

@@

Building upon the previous literature, this study proposes that attitudes toward market justice are shaped by a dynamic interplay between individuals' perceptions of economic inequality and their perceptions of meritocracy. Specifically, we argue that both perceptions independently and interactively influence the extent to which individuals endorse market-based distributions of social goods and services in Chile. To our knowledge, there are not studies so fay analyzing such associations, let alone from a longitudinal perspective.

Firstly, and consistent with previous findings about research on redistributive preferences [@castillo_perception_2022], we expect that a higher perception of economic inequality will be associated with lower market justice preferences. When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay. Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution. This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force in shaping public attitudes toward inequality and justice [@canalesceron_sujeto_2021; @mac-clure_justicia_2024].

@@

Additionally, this study examines whether these relationships vary over time, particularly in light of broader social and political transformations. We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points. While we acknowledge the relevance of contextual events for public opinion and political orientation [@disipavlic_justification_2025], such as the 2019 political outburst or the subsequent constitutional process in CHile, we are aware that identifying causal associations lies outside the scope of this research. Instead, and from an exploratory perspective, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.

### 2025-08-08 resubmission/05-discussion.qmd
jc review

This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile. It posits that higher perceived inequality and stronger meritocratic perceptions both independently contribute to greater support for market-based distributions of social goods. Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic perceptions on market justice preferences. Additionally, it considers how major social and political mobilizations, such as those occurring around 2019 and subsequent years, may have weakened these relationships, reflecting increased critical perspectives on systemic inequalities and market mechanisms. The subsequent discussion interprets the empirical findings in light of these hypotheses, considering their implications for understanding attitudes toward inequality and social justice in a context marked by profound social change.

The first set of hypothesis suggested that perceptions of economic inequality were strongly linked to preferences for market justice. The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021; @garcia-sanchez_vicious_2019].  Such a relationship reflects a moral economy in which perceptions of systemic unfairness undermine the legitimacy of existing distributions and fuel demands for greater equity. In particular, the negative within-individual effect observed over time ($H1_{b}$) suggests that as inequality becomes more salient for some individuals, their support for market justice declines—potentially due to rising distrust in market mechanisms or growing disillusionment with the fairness of the system. However, although the effect of perceived inequality at the between-individual level is also negative, it is not statistically significant. Therefore, we cannot conclude that individuals who, on average, perceive higher inequality are systematically less supportive of market justice. As a result, hypothesis $H1_{a}$ does not receive empirical support.

Regarding the second set of hypotheses, the results confirmed that meritocratic perceptions—particularly those emphasizing individual effort—were associated with stronger support for market-based distribution systems. Individuals who believed that success is primarily achieved through personal effort were more likely to justify unequal access to core social services based on income, interpreting such disparities as outcomes of individual merit rather than systemic injustice ($H2_{a}$). This finding aligns with previous research showing that meritocratic narratives serve as moral justifications that legitimize social stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These perceptions operate symbolically to reinforce structural inequalities by reducing support for redistributive policies, framing inequality as both fair and deserved. Consistent with prior work by @castillo_socialization_2024 on Chilean students, the findings suggest that such meritocratic perceptions uphold existing hierarchies by promoting the acceptance of inequality as a reflection of individual virtue rather than structural failure. This mechanism is particularly salient in neoliberal contexts like Chile, where market logics heavily shape social attitudes [@canalesceron_sujeto_2021].

Interestingly, intra-individual changes in meritocratic perceptions over time ($H2_{b}$) reveal mixed effects. While increases in the belief that rewards are based on individual effort are associated with stronger preferences for market justice, increases in the belief that rewards derive from talent are linked to weaker support for such principles. One possible explanation is that effort is generally viewed as a controllable and malleable trait, aligning with the normative foundations of market fairness. In contrast, talent tends to be perceived as innate and less subject to personal control, rendering talent-based inequality less legitimate. Furthermore, increased exposure to real-world scenarios in which outcomes are clearly shaped by inherent traits rather than hard work may lead individuals to question the fairness and legitimacy of market-based reward systems.

The third set of hypotheses addressed the moderating role of perceptions of inequality in the relationship between meritocratic perceptions and preferences for market justice. The analysis revealed that the positive association between meritocratic perceptions and support for market-based distribution systems tends to be stronger when perceived economic inequality is high—contradicting our initial expectations ($H3_{a}$ and $H3_{b}$). Although this moderating effect is not statistically significant for effort-related meritocratic perceptions, it is significant for talent-related perceptions at both the within- and between-individual levels. This finding runs counter to our hypotheses, as it suggests that the effect of meritocratic perceptions—specifically, talent-based meritocracy—intensifies as perceptions of inequality increase. However, this interpretation requires caution. The main effect of perceived inequality on support for market justice is negative. Thus, as perceived inequality becomes less negative (i.e., closer to zero), the positive relationship between talent-based meritocratic perceptions and market justice preferences becomes stronger. This suggests a more nuanced dynamic: meritocratic perceptions, particularly those emphasizing talent, may serve as a stronger justificatory mechanism for market-based inequalities among individuals who are less inclined to perceive inequality as a problem. In other words, among those who do not strongly perceive systemic disparities, meritocratic narratives may play a more influential role in legitimizing unequal outcomes.

Finally, the fourth hypothesis examined the potential impact of recent social mobilizations—particularly those that erupted in 2019—on the relationship between meritocratic perceptions, perceptions of inequality, and preferences for market justice. The main effect of time indicates that, on average, support for market justice increased following the 2019 protests. However, the analysis did not detect any significant interactions between distributive beliefs and support for market justice, suggesting that the social mobilizations did not substantially modify the underlying associations between these variables. As a result, hypotheses $H4_{a}$ and $H4_{b}$ are not supported by the data. This may point to a certain stability—or even resilience—in the normative frameworks that guide individuals’ evaluations of distributive justice, despite the occurrence of major collective political events. Alternatively, while the protests may have eroded trust in institutional arrangements or governance, they may not have fundamentally altered individuals’ beliefs about how rewards and resources should be allocated in society.

>>>>

The first set of hypotheses suggested that perceptions of economic inequality were strongly linked to preferences for market justice. The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021; @garcia-sanchez_vicious_2019].  In particular, the negative within-individual effect observed over time ($H1_{b}$) suggests that as inequality becomes more salient for some individuals, their support for market justice declines, potentially due to rising distrust in market mechanisms or growing disillusionment with the fairness of the system. However, although the effect of perceived inequality at the between-individual level is also negative, it is not statistically significant. Therefore, we cannot conclude that individuals who, on average, perceive higher inequality are systematically less supportive of market justice. As a result, hypothesis $H1_{a}$ does not receive empirical support.

Regarding the second set of hypotheses, the results confirmed that meritocratic perceptions—particularly those emphasizing individual effort—were associated with stronger support for market-based distribution systems. Individuals who believed that success is primarily achieved through personal effort were more likely to justify unequal access to core social services based on income, interpreting such disparities as outcomes of individual merit rather than systemic injustice ($H2_{a}$). This finding aligns with previous research showing that meritocratic narratives serve as moral justifications that legitimize social stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These perceptions operate symbolically to reinforce structural inequalities by reducing support for redistributive policies, framing inequality as both fair and deserved. Consistent with prior work by @castillo_socialization_2024 on Chilean students, the results suggest that such meritocratic perceptions uphold existing hierarchies by promoting the acceptance of inequality as a reflection of individual virtue rather than structural failure. This mechanism is particularly salient in neoliberal contexts like Chile, where market logics heavily shape social attitudes [@canalesceron_sujeto_2021].

Interestingly, intra-individual changes in meritocratic perceptions over time ($H2_{b}$) reveal mixed effects. While increases in the belief that rewards are based on individual effort are associated with stronger preferences for market justice, increases in the belief that rewards derive from talent are linked to weaker support for such principles. One possible explanation is that effort is generally viewed as a controllable and malleable trait, whereas talent tends to be perceived as innate and less subject to personal control, rendering talent-based inequality less legitimate. Furthermore, increased exposure to real-world scenarios in which outcomes are clearly shaped by inherent traits rather than hard work may lead individuals to question the fairness and legitimacy of market-based reward systems.

The third set of hypotheses addressed the moderating role of perceptions of inequality in the relationship between meritocratic perceptions and preferences for market justice. The analysis revealed that the positive association between meritocratic perceptions and support for market-based distribution systems tends to be stronger when perceived economic inequality is high, contradicting our initial expectations ($H3_{a}$ and $H3_{b}$). Although this moderating effect is not statistically significant for effort-related meritocratic perceptions, it is significant for talent-related perceptions at both the within- and between-individual levels. However, this interpretation requires caution: the main effect of perceived inequality on support for market justice is negative, meaning that as perceived inequality becomes less negative (i.e., closer to zero), the positive relationship between talent-based meritocratic perceptions and market justice preferences becomes stronger. This suggests a more nuanced dynamic: meritocratic perceptions, particularly those emphasizing talent, may serve as a stronger justificatory mechanism for market-based inequalities among individuals who are less inclined to perceive inequality as a problem. In other words, among those who do not strongly perceive systemic disparities, meritocratic narratives may play a more influential role in legitimizing unequal outcomes.

Finally, the fourth exploratory hypothesis examined the potential impact of social events — particularly those that erupted in 2019 — on the relationship between meritocratic perceptions, perceptions of inequality, and preferences for market justice. The main effect of time indicates that, on average, support for market justice increased after 2019 protests. However, the analysis did not detect any significant interactions between distributive beliefs and support for market justice, suggesting that events in this period (as the mobilizations, the pandemic and/or the constitucional processes) did not substantially modify the underlying associations between these variables. As a result, hypotheses $H4_{a}$ and $H4_{b}$ are not supported by the analysis. This may point to a certain stability or even resilience in the normative frameworks that guide individuals’ evaluations of distributive justice, despite the occurrence of major collective political events. Alternatively, while the protests may have eroded trust in institutional arrangements or governance, they may not have fundamentally altered individuals’ beliefs about how rewards and resources should be allocated in society.

### 2025-08-08 resubmission/06-conclusion.qmd
jc review

This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey. By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal and commoditized environment.

The findings confirm that higher perceptions of economic inequality are associated with less support for market justice preferences, this is, the belief that it is fair that those with higher income have better social services such as education, pensions and health. At the same time, meritocratic perceptions—particularly those emphasizing individual effort—are strongly associated with support for market-based distributions, suggesting that meritocracy serve as a moral justification for structural inequalities. However, changes in meritocratic perceptions over time reveal a more nuanced picture: while increased emphasis on effort reinforces support for market justice, increased emphasis on talent tends to reduce it. This distinction likely reflects broader beliefs about the controllability and fairness of different meritocratic traits. Moreover, the interaction between inequality perceptions and meritocratic perceptions indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative—that is, when individuals are less aware or concerned about economic disparities. This finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.

This research advances the extant literature by integrating subjective perceptions with social and political contexts to explain attitudes toward economic inequality and distributional justice. While previous studies primarily focused on objective measures or individual characteristics [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015], this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time. It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness. The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual perceptions and preferences.

Despite its contributions, the study has several limitations. First, the reliance on self-reported perceptions and attitudes may be subject to social desirability bias. Second, the focus on Chile, while offering valuable insights into a specific context, limits the generalizability of results to other countries with different institutional and cultural backgrounds. Third, the observational nature of the data prevents causal inferences, and unobserved confounding factors may influence the observed relationships. Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed.

Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs. Comparative studies across different national contexts would help determine the universality or specificity of these dynamics. Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes. Finally, examining how these perceptions influence behavioral outcomes, such as political participation or support for social movements, would provide valuable insights into the pathways from beliefs to collective action and policy change.

>>>>

This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey. By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal and commodified environment.

The findings support that higher perceptions of economic inequality are associated with less support for market justice preferences, this is, the belief that it is fair that those with higher income have better social services such as education, pensions and health. At the same time, meritocratic perceptions — particularly those emphasizing individual effort — are strongly associated with support for market-based distributions, suggesting that meritocracy serve as a moral justification for structural inequalities. However, changes in meritocratic perceptions over time reveal a more nuanced picture: while increased emphasis on effort reinforces support for market justice, increased emphasis on talent tends to reduce it. This distinction likely reflects broader beliefs about the controllability and fairness of different meritocratic traits. Moreover, the interaction between inequality perceptions and meritocratic perceptions indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative — that is, when individuals are less aware or concerned about economic disparities. Such finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.

This research advances the extant literature by integrating subjective perceptions with social and political contexts to explain attitudes toward economic inequality and distributional justice. While previous studies primarily focused on objective measures or individual characteristics [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015], this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time. Furthermore, it highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness. The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual perceptions and preferences.

Regarding avenues for future research, experimental evidence could help expand the understanding of the association between perceptual variables and market justice preferences. International comparative studies are also necessary in order to assess the role of national contexts and the universality or specificity of these dynamics. Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes and their variation over time. Finally, examining how these perceptions influence behavioral outcomes, such as political participation or support for social movements, would provide valuable insights into the pathways from beliefs to collective action and policy change.
